A Partitioning Algorithm for Maximum Common (Connected) Subgraphs\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/K503058/1 and EP/M508056/1]}

### 2017-02-10 paper/paper.tex
Worked on algorithm description

% TODO: use operatorname for these
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\N}{N}

@@

\newcommand{\examplexG}[5] {
    \begin{minipage}{.2\textwidth}
    \tikz {
        \graph [nodes={draw, circle, minimum width=.6cm}, circular placement, radius=1cm,
                clockwise=5] {
                    1[label=90:#1],2[label=0:#2],3[label=0:#3],4[label=180:#4],5[label=180:#5];
            1--4; 1--5; 2--3; 2--5; 3--5;
        };
    }
    \end{minipage}
}
\newcommand{\examplexH}[6] {
    \begin{minipage}{.2\textwidth}
    \tikz {
        \graph [nodes={draw, circle, minimum width=.6cm}, circular placement, radius=1.1cm,
                clockwise=6, phase=60] {
                    a[label=0:#1],b[label=0:#2],c[label=0:#3],d[label=180:#4],e[label=180:#5],f[label=180:#6];
            a--b; a--c; a--e; b--d; b--f; c--d; c--e; c--f; d--f; e--f;
        };
    }
    \end{minipage}
}

@@

MCS is the problem of finding a maximum-cardinality set of pairs $M = \{(v_1,
w_1), \dots, (v_{|M|}, w_{|M|})\}$, with $v_i \in V(G)$ and $v_i \in V(H)$ for
all $i$, such that for each pair $i,j$ ($1 \leq i < j \leq |M|$), $v_i$ and
$v_j$ are adjacent in $G$ if and only if $w_i$ and $w_j$ are adjacent in $H$.

Let $N(v, G)$ denote the neighbourhood of a vertex $v$ in a graph $G$.  We
extend this notation to denote the set of vertices that are adjacent to all
vertices in a set $U$: $N(U, G) = \bigcap_{v \in U} N_G(v)$.

@@

The most successful existing approaches to solving MCS are reformulation to the
maximum clique problem on one hand, and constraint programming
\cite{DBLP:conf/cp/McCreeshNPS16} on the other. In the clique
encoding, an \textit{association graph} is created with one node for each
pair in the set $\{(u,v) \mid u \in \V(G), v \in \V(H) \}$ and edges representing
pairs of mappings that may be chosen together. A maximum clique in this graph
represents a maximum common subgraph of $G$ and $H$.

In the constraint programming approach, we have a variable corresponding to each
vertex in $G$, and each variable initially has a value in its domain for each vertex
in $H$ to which it may be mapped, plus an additional value $\bot$ signifying that
the vertex remains unmapped. A backtracking search is performed in which values
are assigned to variables, and the domains of unassigned variables are pruned.

In this paper, we introduce a
new branch and bound algorithm for MCS which does not require a domain to be
explicity maintained in memory for each vertex in $\V(G)$. The algorithm
maintains labellings of the unmatched vertices in $\V(G)$ and $\V(H)$ such that
a pair $(v, w)$ with $v \in \V(G)$ and $w \in \V(H)$ can be added to the
mapping if and only if $v$ and $w$ have the same label.

\section{Walkthrough}

We will use the graphs $G$ and $H$ in Figure~\ref{fig:alg1} to demonstrate the
algorithm.  These graphs have a maximum common subgraph with four vertices; the
mapping $\{1a, 2f, 3d, 5b\}$ is one example.

@@

The algorithm proceeds by building up a mapping $M$, starting with the empty
mapping $\{\}$. Begin by (arbitrarily) mapping vertex $1$ in $G$ to vertex $a$
in $H$, giving $M=\{1a\}$.  Label each unmatched vertex according to whether it
is adjacent to $1$ (in $G$) or $a$ (in $H$), as shown in Figure~\ref{fig:alg2}.
Adjacent vertices have label 1; non-adjacent vertices have label 0.  It is
straightforward to see that we can extend $M$ with a vertex mapping $vw$, with
$v \in \V(G)$ and $w \in \V(H)$, if and only if $v$ and $w$ have the same label.  This
property is the algorithm's key invariant.

@@

Observe that our invariant is maintained: we can extend $M$ by a vertex pairing if
and only if the two vertices have the same colour.

@@

    $\{1a, 2d\}$
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $G$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $3$ & 01 \\
        $4$ & 10 \\
        $5$ & 11 \\
    \hline
    \end{tabular}
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $H$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $b$ & 11 \\
        $c$ & 11 \\
        $e$ & 10 \\
        $f$ & 01 \\
    \hline
    \end{tabular}
\end{minipage}
\caption{Step 3}
\label{fig:alg3}
\end{figure}

The algorithm proceeds by backtracking search. When we have explored the full
search space of matchings containing $\{1a, 2d\}$, we try reassigning $2$ to
$f$.  Since $d$ and $f$ are the only vertices to which $2$ can be matched given
the decision to match $1$ to $a$, we lastly explore the possibility that $2$ is
left unmatched, by giving $2$ the label $\bot$ and selecting another vertex in
$\V(G)$ to assign.

\section{The Lilybank Algorithm}

\begin{algorithm}
\DontPrintSemicolon
\nl $expand(\mathit{future},matching)$ \;
\nl \Begin{
\nl \lIf {$future = \emptyset$ \bf{and} $|matching|  > |incumbent|$}{$save(matching)$}
\nl \lIf {$future = \emptyset$}{return}
\nl $bestFuture \gets 0$ \;
\nl \lFor {$\langle P,T \rangle \in future$}{$bestFuture \gets bestFuture + min(|P|,|T|)$}
\nl \lIf {$|matching|  + bestFuture \leq |incumbent|$}{return}
\bigskip
\nl $\langle P,T \rangle \gets select(future)$ \;
\nl $v \gets select(P)$ \;
\nl \For {$w \in T$}{
\nl    $matching \gets matching + (v,w)$\;
\nl    \bf{Let} $future^{'} \gets \emptyset$ \;
\nl    \For {$\langle P^{'},T'^{'}\rangle \in future$}{
\nl        \bf{Let} $P^{''} \gets P^{'} \cap N(v,G) \setminus \{v\}$ \;
\nl        \bf{Let} $T^{''} \gets T^{'} \cap N(w,H) \setminus \{w\}$ \;
\nl        \lIf {$P^{''} \neq \emptyset$ \bf{and} $T^{''} \neq \emptyset$}{$future^{'} \gets future^{'} + \langle P^{''} , T^{''} \rangle$}
\nl        \bf{Let} $P^{''} \gets P^{'} \cap N(v,\overline{G}) \setminus \{v\}$ \;
\nl        \bf{Let} $T^{''} \gets T^{'} \cap N(w,\overline{H}) \setminus \{w\}$ \;
\nl        \lIf {$P^{''} \neq \emptyset$ \bf{and} $T^{''} \neq \emptyset$}{$future^{'} \gets future^{'} + \langle P^{''} , T^{''} \rangle$}
       }
\nl   $expand(future^{'},matching)$ \;
\nl   $matching \gets matching \setminus \{(v,w)\}$ \;
  }
\nl $P \gets P \setminus \{v\}$\;
\nl \lIf {$P = \emptyset$} {$future \gets future \setminus \{\langle P,T \rangle \}$}
\nl $expand(future,matching)$ \;
}
\;
\nl $Split(G,H)$ \;
\nl \Begin{
\nl $incumbent \gets \emptyset$ \;
\nl $expand(\{\langle V(G),V(H) \rangle \},\emptyset)$ \;
\nl return $|incubent|$ \;
}
\caption{Lilybank splitting algorithm}
\label{jtAlg}
\end{algorithm}

Now, let's try and explain this. First, assume we have two graph G and H and there are no more vertices in G than in H (think of G as the pattern graph and H as the target).
future is a list of pairs, where the first element of a pair is a set of pattern vertices P and the second is a set of target vertices T. It is assumed that each pattern vertex in P 
might be matched with any (all) of the target vertices in T, or be unmatched. Therefore we start our search via the function LilybankSplit (line 26). This returns an integer, the size of the largest matching.
In line 29 a call is made to expand, where we have as arguments a list of pairs (called future, line 1) and a set of pairs of the form (v,w) where pattern vertex v is matched to target vertex w.

@@

\paragraph{Heuristics} We ran a small experiment comparing lots of heuristics
such as \dots on random instances. The one where you choose a colour class with
as small a max(left size, right size) as possible seemed to work best. ? Why?

\section{Comparison with Existing Algorithms}

@@

\section{Extensions}

@@

If we have vertex labels, we replace ${\langle V(G),V(H) \rangle}$ in line 28 of
the algorithm with a set containing a colour class for each label that appears in
both $\V(G)$ and $\V(H)$. In the remainder of this section, \textit{labels} refers to
edge labels.

@@

In the labelled, undirected case, each adjacency-matrix entry contains the edge
label, or a null entry 0 if no edge exists between the pair of vertices.  we
can also use Algorithm~\ref{labDirAlg}, by letting $L$ be the union of $\{0\}$
with the set of all labels that appear in the input graphs. Since there may be
up to $n_G + n_H$ distinct labels, the loop in Algorithm~\ref{labDirAlg} may
execute up to $n_G + n_H$ times, and this variation of the algorithm therefore
requires $\BigO{n^2}$ time per search node.  It is possible to modify the
algorithms to run in $\BigO{n \log n}$ per search node; we will give a brief
outline of the method. First, run lines 17-19 of Algorithm~\ref{jtAlg} to
create a new label-class of vertices that are not adjacent to $v$ or $w$, and
remove these vertices from $\langle P',T' \rangle$. Next, sort $P'$ and $T'$ in
ascending order of the label on the edge from $v$ or $w$ to each vertex. We can
then create the colour classes corresponding to each edge label by
simulataneously traversing $P'$ and $T'$ from left to right.

@@

>>>>

% TODO: use operatorname for these
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\N}{N}
\DeclareMathOperator{\vtxlabel}{label}

@@

%\newcommand{\examplexG}[5] {
%    \begin{minipage}{.2\textwidth}
%    \tikz {
%        \graph [nodes={draw, circle, minimum width=.6cm}, circular placement, radius=1cm,
%                clockwise=5] {
%                    1[label=90:#1],2[label=0:#2],3[label=0:#3],4[label=180:#4],5[label=180:#5];
%            1--4; 1--5; 2--3; 2--5; 3--5;
%        };
%    }
%    \end{minipage}
%}
%\newcommand{\examplexH}[6] {
%    \begin{minipage}{.2\textwidth}
%    \tikz {
%        \graph [nodes={draw, circle, minimum width=.6cm}, circular placement, radius=1.1cm,
%                clockwise=6, phase=60] {
%                    a[label=0:#1],b[label=0:#2],c[label=0:#3],d[label=180:#4],e[label=180:#5],f[label=180:#6];
%            a--b; a--c; a--e; b--d; b--f; c--d; c--e; c--f; d--f; e--f;
%        };
%    }
%    \end{minipage}
%}

@@

We assume that $G$ and $H$ are unlabelled, undirected graphs without loops
(Section~\ref{sec:extensions} describes how these restrictions may be relaxed).

The vertices and edge sets of a graph $G$ are denoted $\V(G)$ and $\E(G)$.
The set of vertices adjacent to vertex $v$ in graph $G$ is called the
\emph{neighbourhood} of $v$, and denoted $\N(v, G)$.

We use $n_G$ and $n_H$ to denote $|\V(G)|$ and $|\V(H)|$. We will occasionally use
and $n$ in complexity results to denote $n_G + n_H$.

@@

The most successful existing approaches to solving MCS are reformulation to the
maximum clique problem on one hand, and constraint programming
\cite{DBLP:conf/cp/McCreeshNPS16} on the other. In the clique encoding, an
\textit{association graph} is created with one node for each pair in the set
$\{(u,v) \mid u \in \V(G), v \in \V(H) \}$ and and edges for each pair of
mappings that may be chosen together. A maximum clique in this graph represents
a maximum common subgraph of $G$ and $H$.

In the constraint programming approach, we have a variable corresponding to
each vertex in $G$, and each variable initially has one value in its domain for
each vertex in $H$ to which it may be mapped, plus an additional value $\bot$
signifying that the vertex remains unmapped. A backtracking search is performed
in which values are assigned to variables, and propagation rules are used to
remove values from the domains of unassigned variables.

\section{The \McSplit\ algorithm}

We introduce a branch and bound algorithm, \McSplit, which uses a new
vertex-labelling scheme to avoid explicitly maintaining the set of feasible
mappings for each vertex in $\V(G)$. As we will explore in
Section~\ref{sec:comparison}, the algorithm may be viewed as a version of the
constraint programming forward-checking (CP-FC) algorithm using a compact data
structure that exploits the nature of the MCS problem.  It is sufficiently
different, however, that it is helpful to describe the algorithm without
reference to previous approaches.

%At each node of the search tree, the algorithm
%labels vertices in $\V(G)$ and $\V(H)$ such that a pair $(v, w)$ with $v \in
%\V(G)$ and $w \in \V(H)$ can be added to the mapping if and only if $v$ and $w$
%have the same label.

The \McSplit\ algorithm finds a maximum-cardinality mapping $M^* = \{(v_1, w_1), \dots,
(v_{|M^*|}, w_{|M^*|})\}$, where the $v_i$ are distinct members of $\V(G)$ and the
$w_i$ are distinct members of $\V(H)$, such that $v_i$ and $v_j$ are adjacent
in $G$ if and only if $w_i$ and $w_j$ are adjacent in $H$. It is clear that,
given such a mapping, the subgraph of $G$ induced by $\{v_i, \dots, v_{|M^*|}\}$
and the subgraph of $H$ induced by $\{w_i, \dots, w_{|M^*|}\}$ are a maximum
common subgraph.

We will use the graphs $G$ and $H$ in Figure~\ref{fig:alg1} to demonstrate the
algorithm.  These graphs have a maximum common subgraph with four vertices; the
mapping $\{1a, 2f, 3d, 5b\}$ (where vertex $1$ is assigned to vertex $a$ and so
on) is one example.

@@

The algorithm proceeds by building up a mapping $M$, starting with the empty
mapping $\{\}$. Select a vertex in $\V(G)$ as the first vertex to
be mapped; we will arbitrarily choose vertex $1$. Each of the vertices in $\V(H)$
to which $1$ may be mapped will be tried in turn, and finally the possibility where
$1$ remains unmatched will be tried.

We will begin by mapping vertex $1$ to vertex $a$, giving $M=\{1a\}$.  Label
each unmatched vertex according to whether it is adjacent to $1$ (in $G$) or
$a$ (in $H$), as shown in Figure~\ref{fig:alg2}.  Adjacent vertices have label
1; non-adjacent vertices have label 0.  It is straightforward to see that we
can extend $M$ with a vertex mapping $vw$, with $v \in \V(G)$ and $w \in
\V(H)$, if and only if $v$ and $w$ have the same label.  This property that
two vertices may be mapped together if and only if they share a label is the
algorithm's main invariant.

@@

Observe that our invariant is maintained: we can extend $M$ by a vertex pairing if
and only if the two vertices have the same label. Note also that each label is prefixed
by the label at the previous level of the search tree (Figure~\ref{fig:alg2}).

@@

    $\{1a, 2d\}$
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $G$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $3$ & 01 \\
        $4$ & 10 \\
        $5$ & 11 \\
    \hline
    \end{tabular}
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $H$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $b$ & 11 \\
        $c$ & 11 \\
        $e$ & 10 \\
        $f$ & 01 \\
    \hline
    \end{tabular}
\end{minipage}
\caption{After mapping $2$ to $d$}
\label{fig:alg3}
\end{figure}

The algorithm backtracks when the incumbent best mapping is at least as large
as a calculated bound given $M$ and the current labelling. To demonstrate how
this bound is calculated, we consider the situation one level deeper in the search
tree shown in Figure~\ref{fig:alg4}. Three vertex labels are used: 100, 101, and 111.
The first two of these only appear in one graph, and therefore vertices with these
labels cannot be added to the mapping. The final label, 111, appears once in $G$ and
twice in $H$, and therefore at most one pair with this label can be added to $M$.
Thus, the upper bound on matching size is $4$. In general, for each label $l$
we add to the bound $\min(|\{ v \in \V(G) : \vtxlabel(v)=l\}|,
|\{ v \in \V(H) : \vtxlabel(v)=l \}|)$.

\begin{figure}[ht]
\centering
\begin{minipage}[t]{0.2\linewidth}
    Mapping

    $\{1a, 2d, 3f\}$
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $G$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $4$ & 100 \\
        $5$ & 111 \\
    \hline
    \end{tabular}
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $H$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $b$ & 111 \\
        $c$ & 111 \\
        $e$ & 101 \\
    \hline
    \end{tabular}
\end{minipage}
\caption{After mapping $3$ to $f$}
\label{fig:alg4}
\end{figure}

%When we have explored the full
%search space of matchings containing $\{1a, 2d\}$, we try reassigning $2$ to
%$f$.  Since $d$ and $f$ are the only vertices to which $2$ can be matched given
%the decision to match $1$ to $a$, we lastly explore the possibility that $2$ is
%left unmatched, by giving $2$ the label $\bot$ and selecting another vertex in
%$\V(G)$ to assign.

\paragraph{Efficient representation of label classes} The algorithm as
described so far requires $\BigO{n^2}$ space at each node of the search tree,
since labels may be up to $n_G$ in length. We can reduce this to $\BigO{n}$
by storing a pair $\langle P,T \rangle$ for each label $l$ that is used, where $P$
is the set of vertices in $\V(G)$ labelled $l$, and $T$ is the set of vertices
in $\V(H)$ labelled $l$. Since there are $n_G + n_H$ vertices, at most this number
of label-classes can exist at once, and the total length of the $P$ and $T$ sets
is at most $n_G + n_H$. We can gain further efficiency by discarding during search
any label-class that has an empty $P$ or $T$ set.

\begin{algorithm}
\DontPrintSemicolon
\nl $\FuncSty{expand}(\mathit{future},matching)$ \;
\nl \Begin{
\nl \lIf {$future = \emptyset$ \bf{and} $|matching|  > |incumbent|$}{$save(matching)$}
\nl \lIf {$future = \emptyset$}{return}
\nl $bestFuture \gets 0$ \;
\nl \lFor {$\langle P,T \rangle \in future$}{$bestFuture \gets bestFuture + min(|P|,|T|)$}
\nl \lIf {$|matching|  + bestFuture \leq |incumbent|$}{return}
\bigskip
\nl $\langle P,T \rangle \gets select(future)$ \;
\nl $v \gets select(P)$ \;
\nl \For {$w \in T$}{
\nl    $matching \gets matching + (v,w)$\;
\nl    \bf{Let} $future^{'} \gets \emptyset$ \;
\nl    \For {$\langle P^{'},T'^{'}\rangle \in future$}{
\nl        \bf{Let} $P^{''} \gets P^{'} \cap N(v,G) \setminus \{v\}$ \;
\nl        \bf{Let} $T^{''} \gets T^{'} \cap N(w,H) \setminus \{w\}$ \;
\nl        \lIf {$P^{''} \neq \emptyset$ \bf{and} $T^{''} \neq \emptyset$}{$future^{'} \gets future^{'} + \langle P^{''} , T^{''} \rangle$}
\nl        \bf{Let} $P^{''} \gets P^{'} \cap N(v,\overline{G}) \setminus \{v\}$ \;
\nl        \bf{Let} $T^{''} \gets T^{'} \cap N(w,\overline{H}) \setminus \{w\}$ \;
\nl        \lIf {$P^{''} \neq \emptyset$ \bf{and} $T^{''} \neq \emptyset$}{$future^{'} \gets future^{'} + \langle P^{''} , T^{''} \rangle$}
       }
\nl   $\FuncSty{expand}(future^{'},matching)$ \;
\nl   $matching \gets matching \setminus \{(v,w)\}$ \;
  }
\nl $P \gets P \setminus \{v\}$\;
\nl \lIf {$P = \emptyset$} {$future \gets future \setminus \{\langle P,T \rangle \}$}
\nl $\FuncSty{expand}(future,matching)$ \;
}
\;
\nl $\FuncSty{McSplit}(G,H)$ \;
\nl \Begin{
\nl $incumbent \gets \emptyset$ \;
\nl $\FuncSty{expand}(\{\langle V(G),V(H) \rangle \},\emptyset)$ \;
\nl return $|incubent|$ \;
}
\caption{\McSplit\ algorithm for Maximum Common Subgraph}
\label{jtAlg}
\end{algorithm}

The full algorithm is shown as Algorithm~\ref{jtAlg}, with input graphs $G$ and $H$.
The main recursive procedure, future is a list of pairs, where the first element of a pair
is a set of pattern vertices P and the second is a set of target vertices T. It
is assumed that each pattern vertex in P might be matched with any (all) of the
target vertices in T, or be unmatched. Therefore we start our search via the
function LilybankSplit (line 26). This returns an integer, the size of the
largest matching.  In line 29 a call is made to expand, where we have as
arguments a list of pairs (called future, line 1) and a set of pairs of the
form (v,w) where pattern vertex v is matched to target vertex w.

@@

\paragraph{Heuristics} We ran a small experiment comparing lots of heuristics
such as \dots on random instances. The one where you choose a label class with
as small a max(left size, right size) as possible seemed to work best. ? Why?

\section{Comparison with Existing Algorithms}\label{sec:comparison}

@@

? Mention partition backtracking

\section{Extensions}\label{sec:extensions}

@@

If we have vertex labels, we replace ${\langle V(G),V(H) \rangle}$ in line 28 of
the algorithm with a set containing a label class for each label that appears in
both $\V(G)$ and $\V(H)$. In the remainder of this section, \textit{labels} refers to
edge labels.

@@

In the labelled, undirected case, each adjacency-matrix entry contains the edge
label, or a null entry 0 if no edge exists between the pair of vertices.  we
can also use Algorithm~\ref{labDirAlg}, by letting $L$ be the union of $\{0\}$
with the set of all labels that appear in the input graphs. Since there may be
up to $n_G + n_H$ distinct labels, the loop in Algorithm~\ref{labDirAlg} may
execute up to $n_G + n_H$ times, and this variation of the algorithm therefore
requires $\BigO{n^2}$ time per search node.  It is possible to modify the
algorithms to run in $\BigO{n \log n}$ per search node; we will give a brief
outline of the method. First, run lines 17-19 of Algorithm~\ref{jtAlg} to
create a new label-class of vertices that are not adjacent to $v$ or $w$, and
remove these vertices from $\langle P',T' \rangle$. Next, sort $P'$ and $T'$ in
ascending order of the label on the edge from $v$ or $w$ to each vertex. We can
then create the label classes corresponding to each edge label by
simulataneously traversing $P'$ and $T'$ from left to right.

@@

Application to induced subgraph isomorphism (with almost no changes). Applications
of a similar technique to other problems???
